---
title: "SMML Class 1 Lab"
author: "John Kubale"
date: "8/27/2024"
output: pdf_document
number_sections: yes
fontsize: 12pt
---


```{r setup, include=FALSE, tidy=TRUE}
knitr::opts_chunk$set(echo = TRUE,cache=TRUE, 
                      autodep=TRUE, cache.comments=FALSE,
                      message=FALSE, warning=FALSE)

library(readxl)


```



## I. Made-up income data 1
```{r}
income1<-data.frame("id"=c(1:20),
                          "income_usd"=rnorm(20, mean = 10000, 
                                             sd = 10000/3), 
                          "educ_yr"=rep(12,20))
head(income1)
summary(income1)
```
1. How do income and education look?
2. Can you study the relationship between income and education? Why or why not?


## II. Made-up income data 2
```{r}
income2<-data.frame("id"=c(101:120),
                          "income_usd"=rep(10000, 20), 
                          "educ_yr"=sample(0:16,20,replace=T) )
head(income2)
summary(income2)
```
1. How do income and education look?
2. Can you study the relationship between income and education? Why or why not?


## III. Wage data in R package ISLR2
```{r, echo=FALSE}
library(ISLR2)

```


```{r}
data("Wage")
# run ?Wage in your console to see data description in help
# Can also see p.13 of R manual_ISLR.pdf for data description
dim(Wage)
names(Wage)
head(Wage)
summary(Wage)
```
1. What do you observe?
  + What is the variable type for each variable? 
2. What stories would you like to study using this data?
3. How would you express those stories with formulas? 


## IV. psid data in R package faraway
```{r, echo=FALSE} 
library(faraway)

```

```{r}
data("psid")
# Run ?psid in console to see data description in help
# Can also see p.78 of R manual_faraway.pdf for data description
head(psid)
summary(psid)
```
1. What do you observe?
+ What is the variable type for each variable? 
2. What stories would you like to study using this data?
3. How would you express those stories with formulas? 


## V. Fictitious data 
```{r}
fic_dat7 <- read_excel(
  # "C:/Users/jkubale/Dropbox (University of Michigan)/MPSDS 685/data/Fictitious Data.xlsx",
  "../../data/Fictitious Data.xlsx",
  sheet=7)
head(fic_dat7)

```
- The filepath that has been commented out (line 92) will also work to import the data (assuming that's where the file is located). Why might we prefer to use one way of writing the filepath over the other?


1. Compute the average of Distress_2019 and Distress_2021
```{r}
mean(fic_dat7$Distress_2019)
mean(fic_dat7$Distress_2021)
```
- Can you say that there is a difference in Distress between 2019 and 2022? Why or why not?

2. Compute the average of Distress_2019 and Distress_2021 by Sex
```{r}
cbind(aggregate(Distress_2019~Sex,fic_dat7,FUN=mean),
      aggregate(Distress_2021~Sex,fic_dat7,FUN=mean))
```
- Can you say that there is a difference in Distress between Male and Female in 2019? How about in 2022? Why or why not?


