---
title: "HW 4 Group 1"
author: "Megan Valmidiano & Namit Shrivastava"
date: "2024-09-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. Examine medv as a function of crim, zn and indus in a multiple linear regression.

```{r include=FALSE}
library(MASS)
library(tidyverse)
```

```{r}
#create the model
model <- lm(medv ~ crim + zn + indus, data = Boston)
```

**A. Identify the predictors which are “statistically significant” at α = 0.05.**

```{r}
summary(model)
```

The statistically significant predictors at alpha = 0.05 are the crime, residential zoning, and non-retail business acreage variables.

**B. List the null and alternative hypotheses tested in 1A and your conclusions.**

For the crime variable:
+ Null: There is no relationship between crime and median home value (beta = 0)
+ Alt: There is a relationship between crime and median home value (beta != 0)
+ Conclusion: We can reject the null hypothesis. There is a statistically significant negative relationship between crime and median home value.

For the zone variable:
+ Null: There is no relationship between zone and median home value (beta = 0)
+ Alt: There is a relationship between zone and median home value (beta != 0)
+ Conclusion: We can reject the null hypothesis. There is a statistically significant positive relationship between zone and median home value.

For the industry variable:
+ Null: There is no relationship between industry acreage and median home value (beta = 0)
+ Alt: There is a relationship between industry acreage and median home value (beta != 0)
+ Conclusion: We can reject the null hypothesis. There is a statistically significant negative relationship between industry acreage and median home value.