---
title: "HW 4 Group 1"
author: "Megan Valmidiano & Namit Shrivastava"
date: "2024-09-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. Examine medv as a function of crim, zn and indus in a multiple linear regression.

```{r include=FALSE}
library(MASS)
library(tidyverse)
```

```{r}
#create the model
model <- lm(medv ~ crim + zn + indus, data = Boston)
```

**A. Identify the predictors which are “statistically significant” at α = 0.05.**

```{r}
summary(model)
```

The statistically significant predictors at alpha = 0.05 are the crime, residential zoning, and non-retail business acreage variables.

**B. List the null and alternative hypotheses tested in 1A and your conclusions.**

For the crime variable:
+ Null: There is no relationship between crime and median home value (beta = 0)
+ Alt: There is a relationship between crime and median home value (beta != 0)
+ Conclusion: We can reject the null hypothesis. There is a statistically significant negative relationship between crime and median home value.

For the zone variable:
+ Null: There is no relationship between zone and median home value (beta = 0)
+ Alt: There is a relationship between zone and median home value (beta != 0)
+ Conclusion: We can reject the null hypothesis. There is a statistically significant positive relationship between zone and median home value.

For the industry variable:
+ Null: There is no relationship between industry acreage and median home value (beta = 0)
+ Alt: There is a relationship between industry acreage and median home value (beta != 0)
+ Conclusion: We can reject the null hypothesis. There is a statistically significant negative relationship between industry acreage and median home value.

**C. Interpret each of the regression coefficients as if it were the primary exposure of interest. Do they make sense?**

Crime: As the per capita crime rate increases by one unit, the median home value in a town will decrease by about $248. 

Zone: As the proportion of residential land zoned for lots over 25,000 sq ft increases by one unit, the median home value in a town will increase by about $58.50.

Industry: As the proportion of non-retail business acres increases by one unit, the median home value in a town will decrease by about $415.

**D. It’s generally not good practice to interpret all predictors as if they were the exposure of interest. Why do you think doing so could be problematic?**

In a multivariate model, not all variables are meant to be causal factors. Some variables included in a model are meant to be control variables, or may be confounders. Additionally, interpreting the individual coeficients in a multivariate model may be misleading. If predictors are correlated with one another, such as education and age, their "real" influence is distorted.

**E. Construct and interpret 95% confidence intervals for ˆβcrim, ˆβzn, and ˆβindus (you do not need to calculate them “by hand”). How does the confidence intervals correspond to the hypotheses tested in 1A and 1B?**

```{r}
confint(model, level = 0.95)
```

+ For the crime variable: The confidence interval tells us that there is a 95% chance that the true beta coeficient is between -0.335 and -0.162. This means that our model's estimate is statistically significant, and we can reject the null hypothesis because the interval does not include 0. Additionally, the bounds are both negative so we can say that crime has a negative effect on median home value.
+ For the zone variable: There is a 95% chance that the true beta is betwee 0.024 and 0.093. Our model's estimate is statistically significant, so we can reject the null. This estimate is also precise, as the interval is very small. This variable has a positive effect on median home value.
+ For the industry variable, there is a 95% chance that the true beta is between -0.541 and -0.290. This interval is also sstatistically significant, and we can reject the null hypothesis. There is a negative effect between this variable and median home value.

