---
title: "HW2 Group 1"
author: "Megan Valmidiano & Namit Shrivastava"
date: "2024-09-10"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. 

**Using the Auto dataset in the ISLR2 package fit a regression model of miles per gallon (mpg) as a function of acceleration and save as an object called “mod1”.**

```{r}
#call the ISLR2 package
library(ISLR2)

#librarying the package did not load the data into my environment
#I am creating an object to store my data
#examine the data
head(Auto)

#create a regression model
mod1 <- lm(mpg ~ acceleration, Auto)
```

### 2. 

**How would you interpret the intercept and slope from #1?**

```{r}
summary(mod1)
```

I would interpret the intercept by saying that if a car takes zero seconds to accelerate from 0 to 60 mph, it will have about 4.8 miles per gallon. The slop for acceleration means that for each 1 second increase in the time it takes a car to accelerate from 0 to 60 mph, it will gain about 1.2 miles per gallon.

### 3. 

**Does the intercept make sense? Show how you would refit the model to get an interpretable intercept and explain whether this is necessary. Interpret the intercept from the refit model.**

### 4. 

**What is the null hypothesis being tested via the t statistic for acceleration? Does this hypothesis change if the predictor is centered?**

### 5. 

**What are some of the reasons you might choose a more flexible modeling approach? What are some of the tradeoffs of choosing a more flexible approach?**
