---
title: "HW2 Group 1"
author: "Megan Valmidiano & Namit Shrivastava"
date: "2024-09-10"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. 

**Using the Auto dataset in the ISLR2 package fit a regression model of miles per gallon (mpg) as a function of acceleration and save as an object called “mod1”.**

```{r}
#call the ISLR2 package
library(ISLR2)
#examine the data
head(Auto)

#create a regression model
mod1 <- lm(mpg ~ acceleration, Auto)
```

### 2. 

**How would you interpret the intercept and slope from #1?**

```{r}
summary(mod1)
```

I would interpret the intercept by saying that if a car takes zero seconds to accelerate from 0 to 60 mph, it will have about 4.8 miles per gallon. 
The slope for acceleration means that for each 1 second increase in the time it takes a car to accelerate from 0 to 60 mph, it will gain about 1.2 miles per gallon.

### 3. 

**Does the intercept make sense? Show how you would refit the model to get an interpretable intercept and explain whether this is necessary. Interpret the intercept from the refit model.**

```{r}
Auto$acceleration_centered <- Auto$acceleration - mean(Auto$acceleration)
mod1_centered <- lm(mpg ~ acceleration_centered, data = Auto)
summary(mod1_centered)
```

So a 0 acceleration is not realistic and so conclusively the intercept won't make sense too. Now to get an interpretable
intercept, we can try centering this predictor variable which is acceleration in this scenario.

Further, we can interpret the intercept in the refit model as it represents the expected value of mpg, given that acceleration is 
at its mean value. Now we can say that its more interpretable because it corresponds to a realistic scenario.

### 4. 

**What is the null hypothesis being tested via the t statistic for acceleration? Does this hypothesis change if the predictor is centered?**

The null hypothesis being tested for is that acceleration is not a predictor of miles per gallon. Centering the predictor does not change the relationship between the variables of interest, it just transforms the graph to make the intercept more interpretable. As a result, the null hypothesis that is being tested has not changed with the centered model.

### 5. 

**What are some of the reasons you might choose a more flexible modeling approach? What are some of the tradeoffs of choosing a more flexible approach?**

Ok so, in real world, we have more non-linear relationships to handle. Hence we need a more flexible modelling approach.
Also another advantage being a higher predictive accuracy.

But with more flexibility, the model is prone to becoming more complex and hence might lead to overfitting. Not only that,
but more data and computational resources will be used and sometimes in such a scenario, it might be harder to interpret than
simple linear models.
