---
title: "HW3 Group 1"
author: "Megan Valmidiano & Namit Shrivastava"
date: "2024-09-17"
output: pdf_document
---

This exercise involves the Boston housing data set. Assume that we are interested in per
capita crime rate, crim.

```{r}
library(MASS)
library(tidyverse)
library(ggplot2)
data(Boston)
```
So before we start examining, we load the required libraries and load the Boston dataset

A. Examine crim with summary() and in a histogram.
```{r}
summary(Boston$crim)
ggplot(Boston, aes(x = crim)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  labs(title = "Histogram of Crime Rate (crim)",
       x = "Crime Rate", y = "Frequency")
```
The histogram shows that the distribution of crime rates is heavily right-skewed, 
with most suburbs in Boston having low crime rates and a few suburbs with very high crime rates.

B. Focus on suburbs with the crime rate above 25. How many suburbs fall into this group?
What are the pupil-teacher ratios like in those suburbs? How about property tax rates? How
about median home values? How do the pupil-teacher ratios, property tax rates and median
home values compare between these suburbs and the remaining suburbs?
```{r}
high_crime_suburbs <- Boston %>% filter(crim > 25)
```
```{r}
high_crime_suburbs <- subset(Boston, crim > 25)
```
C. Create a scatter plot of the crime rates and the median home values for 1) all suburbs, 2)
suburbs bounding Charles River, and 3) suburbs not bounding Charles River. What do you
observe?
D. Analyze the crime rates as a function of median home values in a simple linear regression
with an intercept. Report what the regression coefficients mean in lay terms.
E. Calculate the coefficients reported in D as well as their standard errors by hand.
F. Create a scatter plot of the crime rates and the median home values with a regression line.
Is the regression line a good summary of the crime rates? Examine residuals to assess this.
G. Create a scatter plot of predicted crim (x-axis) and residuals (y-axis). What do you
observe?
