---
title: "Homework 5_Megan Valmidiano & Namit Shrivastava_Group1"
format: pdf
editor: visual
---
This exercise involves the Boston housing dataset in ISLR2. Assume that we are interested
in median home values, medv.

1. Examine medv as a function of chas in a simple linear regression model with an intercept.
Notice the nature of chas for its use in this and following models. What hypothesis are you
testing with each coefficient in lay terms? Given the results of the hypothesis testing, what
do the coefficients mean?


So first let's load the necessary libraries and the Boston dataset
```{r}
library(ISLR2)
library(ggplot2)
library(dplyr)

data("Boston")
head(Boston)
```

So now, time to fit medv as a function of chas in a simple linear regression model with an intercept.
```{r}
# Fitting the simple linear regression model
modelSLR <- lm(medv ~ chas, data = Boston)

# Now to display summary of the model
summary(modelSLR)

# Now based on chas category
#we can  also show the mean of medv
Boston %>%
  group_by(chas) %>%
  summarise(mean_medv = mean(medv))
```
Right, so from the output, we can say that the variable 'chas' is binary, indicating whether a 
tract borders the Charles River (which can be said as 1) or not (0).

Now in layman terms, we can test two hypotheses:

* For the Intercept(B_0), which is the expected median home value when, chas = 0 (not bordering Charles River).

* Now as for the chas variable (B_1), we are testing if there is a significant difference in median home values 
for homes that border the Charles River versus those that do not.

So as per the results obtained, we can interpret the results as :

* The intercept value of B_0 = 22.0938, represents the expected median home value (in $1000) for tracts not bordering the Charles River.

* The coefficient of chas (B_1= 6.346) indicates that tracts bordering the Charles River have, on average, 
$6346 higher median home values compared to those that don't.

Finally also seeing the p-value for both the coefficients, we can say both of them are statistically significant since p < 0.05, 
which suggests that the tracts bordering the Charles River have, on average, higher median home values. So there is a real relationship
between chas and medv.

2. Examine medv as a function of chas and indus in a multiple linear regression. What do the
coefficients mean?

So now, time to fit medv as a function of chas and indus in a multiple linear regression.
```{r}
# Fitting the multiple linear regression model
modelMLR <- lm(medv ~ chas + indus, data = Boston)

# summary statistics of the model
summary(modelMLR)
```
So based on the results, we can interpret the coefficients as:

* The intercept (B_0) which has a value of 29.43170 represents the expected median home value when chas = 0 and indus = 0.

* The chas coefficient that is B_1 in this case has a value of 7.47818, meaning, on average, there is an
increase of $7478 in median home value for tracts bordering the Charles River, holding indus constant.

* Now the indus coefficient (B_2) has a value -0.66592, meaning on an average, there is a decrease of $ 665.92 in median 
home value for each unit increase in the proportion of non-retail business acres, holding chas constant.

3. Given the results from #1 and #2, what do you conclude about chas and indus in relation
to medv? Support your conclusion with a visualization.

4. You think that the proportion of non-retail business acres may have a different impact on
the expected average median home values in towns bordering the Charles River vs. those that
don’t. Fit a linear regression model with an interaction term to assess this. What do you
observe? Support your observations with a visualization.

