---
title: "Homework 6_HW3 17_Yael Dejene Beshaw & Namit Shrivastava"
format: pdf
editor: visual
---

# A. Predict the amount that a male with average values (given these data) for status, income, and verbal would spend on gambling along with an appropriate 95% interval.
```{r}
#First we load the necessary libraries
library(faraway)
library(dplyr)

#After that, loading the teengamb dataset
data(teengamb)
head(teengamb)

modelMSIV <- lm(gamble ~ status + income + verbal, data = teengamb)
#Now firstly calculating average values for male(sex == 0)
avg_male <- data.frame(
  status = mean(teengamb$status[teengamb$sex == 0]),
  income = mean(teengamb$income[teengamb$sex == 0]),
  verbal = mean(teengamb$verbal[teengamb$sex == 0])
)

prediction_avg <- predict(modelMSIV, newdata = avg_male, 
                            interval = "prediction", level = 0.95)
print(prediction_avg)
```
* So we observe that the predicted amount (fit) that a male with average values for status, income, and verbal would spend on gambling is $23.74285.

* The lower bound (lwr) of the 95% prediction interval is -$25.86432. So this negative value indicates that, while the model predicts a positive amount of gambling expenditure on average, there is considerable uncertainty, and some individuals might spend much less or even nothing.

* The upper bound (upr) of the 95% prediction interval is $73.35002. This indicates that, while the average prediction is $23.74285, some individuals might spend significantly more on gambling.

* Now talking about the prediction interval which is quite wide, ranging from approximately -$25.86432 to $73.35002. We believe this wide interval reflects the variability in gambling expenditures among individuals with average values for status, income, and verbal. Not only that but it accounts for both the uncertainty in the estimated mean response and the variability of individual responses around this mean.

# B. Repeat the prediction for a male with maximal values (for these data) of status, income, and verbal. Which interval is wider and why is this result expected?
```{r}
predict(modelMSIV, newdata = data.frame(status = max(teengamb$status[teengamb$sex == 0]), 
        income = max(teengamb$income[teengamb$sex == 0]), 
        verbal = max(teengamb$verbal[teengamb$sex == 0])), interval = "prediction")
```

* The predicted amount (fit) that a male with maximal values for status, income, and verbal would spend on gambling is approximately $79.30022.

* The lower bound (lwr) of the 95% prediction interval is $21.6408. This means that we can be 95% confident that the actual gambling expenditure will not be lower than $21.6408.

* The upper bound (upr) of the 95% prediction interval is $136.9597. This means that we can be 95% confident that the actual gambling expenditure will not exceed $136.9597.

So we can say that the interval ranges from approximately $21.6408 to $136.9597. This interval accounts for the variability and uncertainty around the predicted amount for an individual case. Now when we compare the intervals from the previous parts:

* Interval for Average Values:
Predicted amount: $23.74285
95% prediction interval: -$25.86432 to $73.35002

* Interval for Maximal Values:
Predicted amount: $79.30022
95% confidence interval: $21.6408 to $136.9597

So the interval for maximal values is wider than the interval for average values. Now the interval for maximal values is wider because the maximal values are likely to be further from the mean, increasing the variability in the prediction. So when predicting for extreme values, the model's uncertainty increases, leading to a broader prediction interval.

# C. Fit a model with sqrt(gamble) as the outcome but with the same predictors. Now predict the response and estimate the appropriate 95% interval for the individual in #A. Take care to give your answer on the original scale of the response (i.e., gamble not sqrt(gamble)).
```{r}
modelsqrt <- lm(sqrt(gamble) ~ status + income + verbal, data = teengamb)
predict(modelsqrt, newdata = data.frame(status = mean(teengamb$status[teengamb$sex == 0]), 
                income = mean(teengamb$income[teengamb$sex == 0]), 
                verbal = mean(teengamb$verbal[teengamb$sex == 0])), interval = "prediction", level = 0.95)^2
```

* The predicted amount (fit) that a male with average values for status, income, and verbal would spend on gambling is $14.69264 meaning it is the expected gambling expenditure for a male with average values for status, income, and verbal.

* The lower bound of the 95% prediction interval is $0.5292093 which indicates that we can be 95% confident that the actual gambling expenditure will not be lower than $0.5292093.

* The upper bound of the 95% prediction interval is $70.45358 which now indicates that we can be 95% confident that the actual gambling expenditure will not exceed $70.45358.

Finally, the prediction interval ranges from approximately $0.5292093 to $70.45358. This interval accounts for the variability and uncertainty around the predicted amount for an individual case and reflects the variability in gambling expenditures among individuals with average values for status, income, and verbal.


# D. Repeat the prediction for the model in #C for a female with status=20, income=1, verbal=10. Give your answer on the original scale of the response and comment on the credibility of the result.
```{r}
predict(modelsqrt, newdata = data.frame(status = 20, income = 1, verbal = 10, sex = 1), 
        interval = "prediction", level = 0.95)^2
```
* The predicted amount (fit) that a female with status = 20, income = 1, and verbal = 10 would spend on gambling is $6.120695

* The lower bound of the 95% prediction interval is approximately $58.17663 which indicates that we can be 95% confident that the actual gambling expenditure will not be lower than $58.17663.

* The upper bound of the 95% prediction interval is approximately $7.178931 which indicates that we can be 95% confident that the actual gambling expenditure will not exceed $7.178931.

* The prediction interval ranges from $58.17663 to $7.178931. This interval accounts for the variability and uncertainty around the predicted amount for an individual case and reflects the variability in gambling expenditures among individuals with the specified values for status, income, and verbal.

Now, this result is not credible because the model is based on data from males and for this question is being used to make a prediction for a female. Also, the model does not account for the effect of the sex predictor adequately, leading to potentially unreliable predictions for females. Additionally, the prediction interval seems to have an issue with the lower bound being higher than the upper bound (which does not make sense), so, it possibly indicates a potential error in the model or the transformation process.


# E. Now, go back to #A. Consider how much males with average values (given these data) for status, income, and verbal would spend on gambling along with an appropriate 95% interval. How does this interval compare to the interval in #A? Explain whether this is expected and why.
```{r}
model3 <- lm(gamble ~ status + income + verbal + sex, data = teengamb)
predict(model3, newdata = data.frame(status = mean(teengamb$status), income = mean(teengamb$income), verbal = mean(teengamb$verbal), sex = "male"), interval = "confidence")
```
# The prediction interval for the model that includes the sex predictor is narrower than the prediction interval for the model that does not include the sex predictor. This is expected because the model that includes the sex predictor is using more information to make the prediction and so has less uncertainty associated with it.
